### 2022-04-25 introduction/introduction.tex
more

Though the interest for rigorous and formal reasoning dates back to Aristotle,
it intensified at the end of 19th and the beginning of 20th century
\cite{jf-01-modern-logic}. Researchers in that time were mostly interested in
finding a formal language that is expressive enough to describe complex
mathematical theories, yet intuitive and understandable enough to allow
reasonably simple formal reasoning systems. First-order logic seemed to satisfy
both requirements.  This formalism not only allows one to model simple
logical relations such as ``if it rains, the road is muddy'' (or formally
$\cst{rains} \iimplies \cst{muddy}$), but also to quantify over objects as in ``all
people are mortal'' ($\iforall x.\, \cst{human}(x) \iimplies \cst{mortal}(x)$).

\paragraph{Automatic Theorem Proving}The initial study of first-order logic and the rules to reason about statements in
it was very prolific. In 1930 G\"odel \cite{kg-30-completeness-theorem} showed
that there is a system of inference rules (calculus) such that for any valid
statement a proof of validity can be constructed in this system. Around the same
time, the first computers were leaving the imagination of the engineers and
began automating many complicated processes.  Naturally, the question as to
whether a computer can decide if a first-order statement is valid (i.e., a
theorem) arose. Church \cite{ac-36-fol-undecidable} and Turing
\cite{tm-37-undecidable} answered this question negatively in 1936.

Despite this negative result, the prospect of automatically proving theorems of
first-order logic remained too enticing. In 1960, Davis and Putnam
\cite{dp-60-dp} described a procedure for checking validity of a
first-order formula. As first-order logic is undecidable, this procedure
terminated only on valid formulas. Even though it was efficient enough to prove
only simple formulas, it was an impressive achievement for the time.

@@

We have identified three stops on the road to full higher-order logic. A
distinguished feature of higher-order logic is that it not only supports
quantification over objects, but also over functions that manipulate these
objects. Thus, higher-order logic allows us to make a statement ``all functions
with arguments $0$ and $1$ result in $2$" ($\iforall x. \,  x (\cst{0}, \cst{1}) \eq
\cst{2}$). Replacing $x$ with the addition function ($\cst{+}$) is enough to
disprove this statement. The first stop is to extend E to support
quantification over functions that are already present in the theory.

Higher-order logic also supports expressing functions that are not explicitly
present in the theory. Suppose we want to prove there is a
function that returns 4 when given 2 and 5 when given 3 ($\iexists x. x(\cst{2})
\eq \cst{4} \lland x (\cst{3}) \eq \cst{5} $). An example of such function is
the function that adds 2 to its argument. In higher-order logic such unnamed
functions can be represented using $\lambda$-abstractions as $\lambda y.\, y \,
\cst{+} \, \cst{2}$. This term suffices to prove the original statement. The
second stop is to support automatically synthesizing such functions.

>>>>

Though the interest for rigorous and formal reasoning dates back to Aristotle,
it intensified at the end of 19th and the beginning of 20th century
\cite{jf-01-modern-logic}. Researchers in that time were mostly interested in
finding a formal language that is expressive enough to describe many
mathematical theories, yet intuitive and understandable enough to allow
reasonably simple formal reasoning systems. Arguably, first-order logic seemed to satisfy
both requirements.  This formalism not only allows one to model simple logical
relations such as ``if it rains, the road is muddy'' (or formally $\cst{rains}
\iimplies \cst{muddy}$), but also to quantify over objects as in ``all people
are mortal'' ($\iforall x.\, \cst{human}(x) \iimplies \cst{mortal}(x)$).

\paragraph{Automatic Theorem Proving}The initial study of first-order logic and
the rules to reason about statements in it was prolific. In 1930 G\"odel
\cite{kg-30-completeness-theorem} showed that there is a system of inference
rules (calculus) such that for any valid statement (i.e., a theorem) a proof can be
constructed in this system. Around the same time, the first computers were
leaving the imagination of the engineers and began automating many complicated
processes.  Naturally, the question as to whether a computer can decide if a
first-order statement is valid  arose. Church
\cite{ac-36-fol-undecidable} and Turing \cite{tm-37-undecidable} answered this
question negatively in 1936.

Despite this negative result, the prospect of automatically proving theorems of
first-order logic remained too enticing. In 1960, Davis and Putnam
\cite{dp-60-dp} described a procedure for checking validity of a first-order
formula. This procedure terminated only on valid formulas, as Church and Turing
proved no procedure can check validity of an arbitrary (possibly invalid) formula
in finite time. Even though it was efficient enough to prove only simple
formulas, it was an impressive achievement for the time.

@@

We have identified three stops on the road to full higher-order logic. A
distinguished feature of higher-order logic is that it not only supports
quantification over objects, but also over functions that manipulate these
objects. Thus, higher-order logic allows us to make a statement ``all functions
with arguments $0$ and $1$ result in $2$" ($\iforall x. \,  x (\cst{0}, \cst{1}) \eq
\cst{2}$). Replacing $x$ with the addition function ($\cst{+}$) is enough to
disprove this statement. The first stop is to extend E to support
quantification over functions that are already present in the original problem.

Higher-order logic also supports expressing functions that are not explicitly
present in the original problem. Suppose we want to prove there is a
function that returns 4 when given 2 and 5 when given 3 ($\iexists x. x(\cst{2})
\eq \cst{4} \lland x (\cst{3}) \eq \cst{5} $). An example of such function is
the function that adds 2 to its argument. In higher-order logic such unnamed
functions can be represented using $\lambda$-abstractions as $\lambda y.\, y \,
\cst{+} \, \cst{2}$. This term suffices to prove the original statement. The
second stop is to support automatically synthesizing such functions.

### 2022-04-25 summary/summary.tex
more

In an attempt to automate significant parts of this proving process, assistants
can invoke automatic theorem provers to finish the proof. However, most
assistants are based on higher-order logic, while most automatic theorem provers
are based on first-order logic. This means that the two systems must communicate
through translations, which obfuscate the problem and likely make it more
difficult than it originally was.

@@

The second part concerns adding support for $\lambda$-abstraction. The most
important challenge we faced during this extension is that of higher-order
unification. At the time we started, the state-of-the-art procedure
for full higher-order unification was developed in 1970s. We designed a new
procedure inspired by this one that enumerates fewer redundant unifiers, has many
modern optimizations built-in, and can easily be customized to trade its
completeness for performance. It is implemented in a
prototype prover called Zipperposition, a less efficient but more easily extendible
prover compared to E. Our evaluation shows that our
procedure substantially improves on the state of the art.

The last part concerns adding support for
first-class Boolean terms. For this part we took inspiration from traditional
automatic higher-order provers, based on tableaux. We fitted those techniques in the
superposition context and implemented them in Zipperposition. They
made Zipperposition the best prover in the higher-order division of the annual CASC
theorem prover competition for two consecutive years.

@@

In conclusion, the work described in this thesis shows that first- and
higher-order provers are much more alike than previously thought. Furthermore,
we showed that through carefully designed and tuned extension, first-order
prover can become an award-winning higher-order prover.

>>>>

In an attempt to automate significant parts of this proving process, assistants
can invoke automatic theorem provers to finish the proof. However, most
assistants are based on higher-order logic, while most automatic theorem provers
are based on first-order logic. This means that the two systems must communicate
through translations, which obfuscate the conjecture being proved and likely make it more
difficult than it originally was.

@@

The second part concerns adding support for $\lambda$-abstraction. The most
important challenge we faced during this extension is that of higher-order
unification. At the time we started, the state-of-the-art procedure
for full higher-order unification was the one developed in 1970s. We designed a new
procedure inspired by this one that enumerates fewer redundant unifiers, has many
modern optimizations built-in, and can easily be customized to trade its
completeness for performance. It is implemented in a
prototype prover called Zipperposition, a less efficient but more easily extendible
prover compared to E. Our evaluation shows that our
procedure substantially improves on the state of the art.

The last part concerns adding native support for
Boolean terms. For this part we took inspiration from traditional
automatic higher-order provers, based on tableaux. We fitted those techniques in the
superposition context and implemented them in Zipperposition. They
made Zipperposition the best prover in the higher-order division of the annual CASC
theorem prover competition for two consecutive years.

@@

In conclusion, the work described in this thesis shows that first- and
higher-order provers are much more alike than previously thought. Furthermore,
we showed that through carefully designed and tuned extension, a first-order
prover can become an award-winning higher-order prover.
